A Practical Optional Type System for Clojure

### 2012-10-24 dissertation/design-choices.tex
Hmap design

\begin{verbatim}
HMaps only convey positive information, not negative.
Map concatenation is not approached, see lit review, many
complications. Instead, only support assoc, dissoc, get.
\end{verbatim}

As a key idiomatic feature of Clojure, we treat heterogeneous maps more formally.
This also serves as a prototype for formalising other aspects.
Here we give a formal syntax, operational semantics and type system using inference
rules, as is standard in the programming language community (see, eg., Pierce~\cite{Pie02}).

>>>>

Heterogeneous maps in Typed Clojure are simple enough to avoid
the complications associated with concatenable records (section \ref{sec:recordtypes})
while still supporting many Clojure idioms.
They support adding new entries with keyword keys, removing entries,
and lookup of keyword keys.
When using these operations with non-keyword keys, heterogeneous maps
are treated like plain hash-maps.

They do not support concatenation (merging), as heterogeneous maps
only hold positive information on the keys with entries, not
which keys are \emph{absent}.
This makes concatenation unsafe, as keys to the right of a
concatentation can ``overwrite'' keys to the left. With no information
on the absence of keys, concatenation becomes an unsound operation.

We formalised some aspects of the behaviour of heterogeneous map types in Appendix~\ref{sec:hmaptheory}.
This also serves as a prototype for formalising other aspects in the future.
Here we give a formal syntax, operational semantics and type system using inference
rules, as is standard in the programming language community (see, eg., Pierce~\cite{Pie02}).

### 2012-10-24 dissertation/future-work.tex
Hmap design

@@

It would be desirable to port Typed Clojure to other Clojure implementations.
Each implementation would bring its own set of challenges in

\subsection{Extensions to heterogeneous map types}

\subsection{Length information and intersection types}

@@

\subsection{Linter}

>>>>

\section{Paired Arguments}

It is not clear how to handle several key Clojure functions
that accept an even number of arguments.
For example, valid usages of \lstinline|assoc| are
\lstinline[mathescape]{(assoc m k v $\overrightarrow{k v}$)},
which takes three arguments and any number of paired arguments.

Strickland, Tobin-Hochstadt, and Felleisen's calculus~\cite{STF09}
is insufficient to express this pattern.
Devising and integrating types that can express this pattern 
is set as future work.

@@

It would be desirable to port Typed Clojure to other Clojure implementations.
Each implementation would bring its own set of challenges with interoperability.
The core of Typed Clojure would preserved with each port changing at least 
its interoperability with its host platform. Section \label{sec:portability}
discusses potential problems with the porting process, and how
they might be solved.

%\subsection{Extensions to heterogeneous map types}

@@

%\subsection{Linter}

### 2012-10-24 dissertation/lit-review.tex
Hmap design

This section 
\cite{ACPR94}

@@

% Rowan
% You've missed some of the steps. Most importantly the move to program analysis
% techniques like Set Based Analysis (SBA) - which suffered from modularity issues
% due to the lack of declared intended types.
% Maybe quote Felleisen - "problem: Mr Spidey and friends infer brittle and large
% types; errors remain difficult to explain and fix"

Type systems for Scheme have since moved away from soft typing
to to other approaches like gradual typing.
Alas, the only reasonably complete reference for commentary on this transition 
appears to be slides from a talk by Felleisen~\cite{Fell09},
a leader in this area of research for over 20 years.
Felleisen says that while Soft Scheme discovered type problems, 
it suffers from incomprehensible
error reporting that required PhD-level expertise to decipher.

\begin{verbatim}
  Bidirectionality
\end{verbatim}

@@

Gradual typing combines static and runtime type checking so programmers
can the most appropriate one for the situation.

Typed Racket was developed as a path for module-by-module
porting of existing untyped Racket modules to a typed sister language~\cite{Tob10}.
Once a module is ported and type checked, it is protected from untyped modules
by inserting runtime checks.

\begin{verbatim}
  Different levels of Gradual Typing discussed by Siek..
\end{verbatim}

@@

Typed Racket includes safe interoperability between any combination of typed and untyped 
Racket modules~\cite{Tob10,TF08}. 
Interactions with untyped modules are protected by adding runtime checks based on expected types.
Typed Racket implements a sophisticated blame calculus. It ensures 
error messages correctly \emph{blame} the source of type errors,
which can be difficult to determine in the presence of higher-order functions. 
\begin{verbatim}references, Walder etc.\end{verbatim}

\section{Record Types}

OCaml-style extensible record types has been the subject of extensive research~\cite{Wan89,CM91,HP91}.
Typed Clojure's heterogeneous map types show some resemblance to extensible
record types, but a survey of the literature did not reveal research flexible
enough to capture common usages of maps in Clojure.
This most likely implies that designing such a system for Typed Clojure would be hard.
For this reason, heterogeneous map types are kept as simple as possible while still being useful.

@@

% Hayashi Singleton types
% Intersection types - CDV81
% Unions in soft typing CF91
% - Static Type Inference in a Dynamically Typed Lanugage
%  - cited by CF91, which has unions

% Unions, intersections, and singletons all fit best with extrinsic types
% - this is one reason why they are similar

>>>>

Typed Clojure adds static type checking to an existing untyped language.
Coming from the other direction, Albardi, Cardelli, Pierce and Didier~\cite{ACPR94}
extend a static language to be more dynamic.
This work centres around the type \emph{Dynamic}, which represents
a static type in a dynamically typed languages.

Rossberg~\cite{Ros07} describes the advantages of having a type
\emph{Dynamic} over other approaches to making statically typed languages
more like dynamic ones. He says that type \emph{Dynamic} models
allows aspects of dynamic style to be represented in statically
typed languages without compromising the type system.
However, he claims type \emph{Dynamic} is rarely used in practical languages
because they are too inconvenient for many cases and they destroy
a valuable property of polymorphic, statically typed languages.

@@

Type systems for Scheme have since moved away from soft typing.
Alas, the only reasonably complete reference for commentary on this transition 
appears to be slides from a talk by Felleisen~\cite{Fell09},
a leader in this area of research for over 20 years.
Felleisen says that while Soft Scheme discovered type problems, 
it suffers from incomprehensible
error reporting that required PhD-level expertise to decipher.

\subsection{Program Analysis}

Scheme then moved to program analysis techniques like Set Based Analysis (SBA).
They, however, suffered from modularity issues due to the lack of declared intended types.
MrSpidey\footnote{\url{http://www.plt-scheme.org/software/mrspidey/}}
is a static type checker for Racket that uses uses SBA.
Felleisen states the main problem with these kinds of systems: ``Mr Spidey and friends infer brittle and large
types; errors remain difficult to explain and fix''~\cite{Fell09}.

@@

Gradual typing combines static and runtime type checking so programmers
can the most appropriate one for the situation.
A key feature of gradual type systems is its use of \emph{bidirectional checking}
like~\cite{PT00}, which requires type annotations in natural places like functions.
Type errors in gradual type systems are often more manageable than in Soft Typing and SBA,
often more comparable to type errors in statically typed languages.

Gradually typed languages have different degrees of runtime checking.
Typed Racket was developed as a path for module-by-module
porting of existing untyped Racket modules to a typed sister language~\cite{Tob10}.
Once a module is ported and type checked, it is protected from untyped modules
by inserting runtime checks.
Typescript~\cite{TypeS2012}, a gradual type system for Javascript, 
does not add runtime checks to untyped interactions.
Runtime checks affect performance, but give better errors and ensure
type errors result in the program failing quicker.

@@

Typed Racket includes safe interoperability between any combination of typed and untyped 
Racket modules~\cite{Tob10,TF08}. 
Interactions with untyped modules are protected by adding runtime checks based on expected types.
Typed Racket implements a sophisticated blame calculus, based on Wadler~\cite{WF09}. It ensures 
error messages correctly \emph{blame} the source of type errors,
which can be difficult to determine in the presence of higher-order functions.

\section{Record Types}
\label{sec:recordtypes}

O'Caml-style extensible record types has been the subject of extensive research 
(eg. Wand~\cite{Wan89}, Cardelli and Mitchell~\cite{CM91}, Harper and Pierce~\cite{HP91}).
Typed Clojure's heterogeneous map types show some resemblance to extensible
record types, but a survey of the literature did not reveal research flexible
enough to capture common usages of maps in Clojure.
This most likely implies that designing such a system for Typed Clojure would be hard.
For this reason, heterogeneous map types are kept as simple as possible while still being useful.

@@
